BOLD Parameter Estimation Using Particle Filters

### 2010-06-30 doc/thesis/particlefilters.tex
close to finished with the particlefilters

\begin{algorithm}
\caption{Stratified Resampling Algorithm}
\begin{algorithmic}

\end{algorithmic}
\end{algorithm}

Unfortunately, this isn't necessarily the truth: since the support is
still limited to the original particles, the number of unique particles can only go down.
This effect, often dubbed "particle impoverishment" can result in excessive quantization
errors in the final distribution. However, there is a solution. Instead of sampling from the
discrete distribution, a smoothing kernel is applied, and $\hat{\chi}_j$ are drawn from
that distribution. Because the distribution is continuous, there is no way for a collapse
of the particles to occur. The question then, is how to decide on the smoothing kernel. 
Often times the easiest way to sample from the continous distribution is to break the 
re-sampling down into two steps. First a member of the discrete distribution is randomly
selected based on the weights, and then based on the smoothing a nearby state variable 
is selected. The process of the selection will be defined as:
\begin{equation}
\chi_i = x_i + h\sigma \epsilon
\end{equation}
Where $h$ is the bandwidth, $\sigma$ is the standard deviation such that $\sigma \sigma^T = cov(x)$
and $\epsilon$ is drawn from the chosen kernel.
It has been proven that when all the elements of the mixture
have the same weight, as is the case after basic resampling, the kernel that minimizes the 
MSE between the estimated and true posterior is the Epanechnikov Kernel (cite Improving Regularised
Particle Filters, C Musso, N Oudjane and F LeGrand). 
\begin{equation}
K = \left\{
\begin{array}{lr}
\frac{n_x+2}{2c_{n_x}}(1-\|x\|^2) & if\ \|x\| < 1\\
0 & otherwise
\end{array}\right.
\end{equation}
%<more here>

>>>>

\begin{algorithm}
\caption{Resampling Algorithm}
\begin{algorithmic}
\STATE Calculate total weight, $W = \sum_{i=0}^{N_p} w^i$
\FORALL{$0 < i < N_p$}
    \STATE Draw $V$ from uniform range $[0, W]$
    \STATE $C = W_t$
    \FORALL{$0 < j < N_p$ and $C < V$}
        \STATE $C = C - w^j$
    \ENDFOR
    \STATE Add $[x^j, \frac{1}{N_p}]$ to the new distribution
\ENDFOR
\STATE 
\end{algorithmic}
\label{alg:Resampling}
\end{algorithm}
For infinite particles this new distribution will match the old.
Unfortunately, this isn't the truth in practice: since the support is
still limited to the original particles, the number of \emph{unique} particles can only go down.
This effect, dubbed "particle impoverishment" can result in excessive quantization
errors in the final distribution. However, there is a solution. Instead of sampling from the
discrete distribution, a smoothing kernel is applied, and particles are drawn from
that distribution. Because the distribution is continuous, particle impoverishment
cannot occur. The easiest way to sample from the continous distribution is to break the 
re-sampling down into two steps. After calculating an estimate of the scale of the original
distribution, \autoref{alg:Resampling} is performed. Next, a distribution is generated
based on the original distributions, although it should have a somewhat smaller variance.
Finally, for each particle in the resampled distribution, a sample is drawn from the smoothing 
distribution and added to the particle.  The process of the selection will be defined as:

\begin{equation}
x_i = x_i + h\sigma \epsilon
\end{equation}

Where $h$ is an optional bandwidth, $\sigma$ is the standard deviation such that 
$\sigma \sigma^T = cov(x)$
and $\epsilon$ is drawn from the chosen kernel. \cite{Musso2001a} goes into
significant 
depth and proves the optimality of the Epanechnikov Kernel for reducing 
MSE between the original and resampled distributions. However, \cite{Musso2001a}
also espouses the usefullness of the Gaussian Kernel, due to the ease
drawing samples from it, which for this work was more important.

\begin{algorithm}
\caption{Regularized Resampling Algorithm}
\begin{algorithmic}
\STATE Calculate Covariance, $C$, of empirical distribution, $\hat{x}$
\STATE Find $D$ such that $DD^T = C$
\STATE Resample $\hat{x}$ using \autoref{alg:Resampling}
\FOR{$0 < i < N_p$}
    \STATE Draw $\epsilon$ from the standard normal, same dimensionality as $X$
    \STATE $x^i = x^i + hD\epsilon$
\ENDFOR
\end{algorithmic}
\label{alg:Resampling}
\end{algorithm}

Thus, with regularized resampling, it is possible to prevent both
particle degeneracy as well as particle impoverishment. An additional
bonus is that as the particle filter converges, the density of particles
in the area of the solution goes up. This has a similar effect to 
simulated annealing where, as the algorithm approaches the end, the
random steps get smaller and smaller. On the down side,
if for some reason the real solution was not covered by the new support 
generated by the resampling algorithm the algorithm could
